---
title: "Gov 50 Final Project"
author: "Gloria Maldonado"
description: My final project
output: distill::distill_article
---



## Project thoughts

The impact of the COVID-19 Pandemic was one that was seen worldwide. As countries, communities, and economies closed down; one of the sectors that was severely impacted was that of education. Schools closed down, many institutions moved to online courses, and students stopped interacting with others as much. The possibility of the pandemic heavily impacting the sector of education was one of my fears. Children in K-12 education may be susceptible to further impact because of their young age and potential hold it may cause to their learning schedules. 
In this project, I hope to analyze if the long-standing debate of Charter versus regular Public Schools. In this project, I am focusing on the Chicago Public Schools Network after having obtained access to their testing and School Records. In these data sets we are able to see the SAT averages of different schools within the CPS district including CPS Network schools which are CPS sponsored Public School and Charter Schools which will be the object of my focus. 
Do Charter Schools perform better than CPS Network Schools? How has Covid-19 impacted school performance through the measurement of SAT Composite Scores and Percentage of College Readiness? 
I hypothesize that Charter schools will outperform CPS Network Schools, and that the Covid 19 Pandemic will have minimal impacts on student performance. I believe this because Charter Schools receive an additional amount of funding from outside donors in addition to receiving federal funds as well; because they were founded to expand opportunities, I feel as if it would make sense to see them outperform the CPS Network Schools which may have less funding. I also believe that the cohort of highschoolers who took the SAT shouldn’t be heaving affected by the change in teaching caused by the Covid19 pandemic because at this point they would have only endured a year of impacted learning.



# Variable Name	Descriptions
"School and Test"	School Year	The school year in which the tests were taken
	School ID	Six digit school identification number
	School Name	School's short name
	Test	Assessment (i.e., PSAT9, PSAT10, SAT)
Attainment	# Students	Number of students taking the PSAT or SAT assessment who received a Total Score. For information on inclusion and exclusion rules, see below.
	Average Composite Score 	The average composite score for the students taking the test.
	# Students with EBRW Score	Number of students taking the PSAT or SAT assessment who received an EBRW Score. For information on inclusion and exclusion rules, see below.
	Average EBRW Score	The average Evidence-Based Reading and Writing score for the students taking the  Spring PSAT or SAT assessment.
	# Students with Math Score	Number of students taking the PSAT or SAT assessment who received a Math Score. For information on inclusion and exclusion rules, see below.
	Average Math Score	The average Math score for the students taking the Spring PSAT or SAT assessment.
	% Meeting College Readiness Benchmark	The percentage of students taking the Spring PSAT or SAT assessment who scored at or above the grade-appropriate combined college readiness benchmark.
	% Partially Met State Standards	The percentage of students who only partially met standards and demonstrated minimal understanding on the PSAT or SAT.
	% Approached State Standards	The percentage of students who were approaching the proficiency level and demonstrated incomplete understanding on the PSAT or SAT.
	% Met State Standards	The percentage of students who met the proficiency level and demonstrated adequate understanding on the PSAT or SAT.
	% Exceeded State Standards	The percentage of students who exceeded the proficiency level and demonstrated a thorough understanding on the PSAT or SAT.
	% Met or Exceeded State Standards	The percentage of students who met or exceeded the proficiency level set by the Illinois State Board of Education for students on the PSAT or SAT.



###The variables that I will be looking at is the Type of School: (CPS Network or Charter) and I will see how they perform on their SAT through their SAT Composite Score. I will also be separating the data so we are able to compare pre-pandemic 2017-2018 alongside 2020-2021.

###In this section I will be analyzing the SAT Composite Scores (dependent variable) and the Type of School (independent variable). This is because I am under the impression that a school with more resources as a Charter will produce students with higher SATs.

```{r}
library(tidyverse)
library(tidyselect)
library(dplyr)


library(readr)
new_sat_data <- read_csv("~/Downloads/Gov 50/new_sat_data.csv")
View(new_sat_data)

library(readr)
type_school_sat <- read_csv("~/Downloads/Gov 50/type_school_sat.csv")

``` 

###Fixing up the Data Set 

```{r}

new_join_sat <- inner_join(new_sat_data, type_school_sat, by = "School ID" )

CPS_SAT<- new_join_sat

CPS_SAT<- CPS_SAT|>
  rename("Percent College Readiness" = contains( "Readiness", ignore.case = TRUE, vars = NULL))|>
  filter(Test == "SAT")


## Making a data set for 2020-2021 School Year 

CPS_SAT_20_21 <- CPS_SAT|> 
  filter(`School Year` == "2020-2021")
```


#Box Plot analysing the 2020-2021 School Year's SAT results by Type of School 
```{r}

#investigating based on 2020-2021 School Year 

CPS_SAT_20_21_plot <- CPS_SAT_20_21|>
  select(`School Year`== "2020-2021")|>
  group_by( `Type of School`)|>
  ggplot( data = CPS_SAT_20_21, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "2020-2021 SAT Averages by Type of School",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )

CPS_SAT_20_21_plot

```



#Calculating the Average Composite Score by Type of School Tibble 
``` {r}
#Average of the `Average Composite Scores` by `Type of School`

avg_sat_table_20_21<- CPS_SAT_20_21 |>
  group_by(`Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)

avg_sat_table_20_21
```




###Is there a difference between CPS Netowrk Schools and Charter Schools? I will be calculating the Average Treatment effect for the 2020-2021 School Year 
```{r}

#Finding out the difference in averages between Charters and CPS Network 
# this is the difference between the SAT SCORE OF CPS and a Charter 

CPS_SAT_20_21|>
  group_by(`Type of School`)|> 
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)

sat_ate <- CPS_SAT_20_21|>
  group_by(`Type of School`)|> 
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`) |> 
  mutate(sat_ate = `Charter` - `CPS Network`)|>
  pull(sat_ate)

sat_ate
```
### When analysing the results of the 2020-2021 Average composite scores by schools we are able to see that the CPS Network is one that has the most range as it has point in the very high almost 1600 area and has some very low scores lower than 800 significantly. In comparison the Charter's avergae is also very similar as they are located around the mid 800s range. It is important to analyze alongside the Tibble, the average composite score for the CPS Network Schools is 872 while the average composite score for the Charter Schools is 855. We are able to see that the CPS Network schools have a higher average. When looking back at our Hypothesis we are seeing that CPS Network may be doing better in SAT Cmposite Scores on average than their Charter Counterparts.When calculating the ATE as the Treatment is going to a Charter School and the control is the CPS Network school.  The ATE shows us that Charter schools on average arescoring -16.32308 less than CPS Network Schoolsr on the SAT.


###Now let's explore another variable:College readiness in the same manner that we did for the previous variable.


#Average Percent of College Readiness by Type of School in the 2020-2021 School Year
```{r} 

college_readiness_average <- CPS_SAT_20_21|> 
  group_by(`Type of School`)|>
  summarise(college_readiness_avg = mean(`Percent College Readiness` , na.rm = TRUE))

college_readiness_average
```

###Is there a benefit of going to a Charter School as opposed to a CPS Network School in the case of College Readiness? ATE Calculation amoung the 2020-2021 School Year for COllege Readiness
```{r}
#We will see if there is an average treatement effect for going to charter 

CPS_SAT_20_21 |>
  group_by(`Type of School`)|>
  summarise(college_readiness_avg = mean(`Percent College Readiness` , na.rm = TRUE))|>
  pivot_wider(values_from = college_readiness_avg, names_from = `Type of School`)

college_readiness_ate<- CPS_SAT_20_21 |>
  group_by(`Type of School`)|>
  summarise(college_readiness_avg = mean(`Percent College Readiness` , na.rm = TRUE))|>
  pivot_wider(values_from = college_readiness_avg, names_from = `Type of School`)|>
  mutate(college_readiness_ate = `Charter` - `CPS Network`)|>
  pull(college_readiness_ate)

college_readiness_ate

```

###After looking at this subsequent variable of College Readiness, which is measured in percents, we are able to see that the average college readiness per type of school varies slightly. It seems that CPS Network schools are on aberage 17.3 percent in College Readiness, while the Charter Schools are 12.5 percent. When calculating the ATE as the Treatment is going to a Charter School and the control is the CPS Network school. The ATE for college readiness is -4.770769, this meaning that on average a Charter student is -4.70769 less college ready than a CPS Network Student. 




### Lets Create a Bootstrap to find out the various quantiles of the averages of different types of Schools in the School year 2020-2021. Lets see what the lower and higher confidence intervals can show us 

```{r}
library(infer)
set.seed(02138) 

ate_boots_20_21<- CPS_SAT_20_21|>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000)|>
  group_by(replicate, `Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg , names_from = `Type of School`)|>
  mutate(sat_ate = `Charter` - `CPS Network`)



ate_sat_95_20_21 <- ate_boots_20_21|>
  ungroup(replicate)|>
  summarize(
    lower_ci = quantile(sat_ate, probs = .025),
    upper_ci = quantile(sat_ate, probs = 0.975)
  )

ate_sat_95_20_21

```

### In this case we are able to see that in the 95% confidence interval we are able to see that on average in most cases a CHarter School will outperform a CPS Network School by 15.7 point more in their SAT Composite score. In the lower confidence interva we are able to see that in some small cases the CPS Network school outperforms the Charter by 46.7 points, reflected in the -46.7 lower_ci. This counters what we have previously seen, showing that in many cases the Charter are indeed outperforming, yet there are  instances in which CPS Network Schools outperform the Charter schools by nearly 46.7 points. The Range of the CPS Network schools seemes to be bigger in this case.




#Lets analyze 2017-2018 school year, will we be able to see a change over the years when looking at the most recent 2020-2021? Lets create another bootstap for the 2017-2018 School Year. What will the upper and lower quantiles tell us? 
```{r}
library(infer)
set.seed(02138) 

CPS_SAT_17_18 <- CPS_SAT|>
  filter(`School Year` == "2017-2018")



ate_boots_17_18<- CPS_SAT_17_18|>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000)|>
  group_by(replicate, `Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg , names_from = `Type of School`)|>
  mutate(sat_ate = `Charter` - `CPS Network`)


ate_sat_95_17_18 <- ate_boots_17_18|>
  ungroup(replicate)|>
  summarize(
    lower_ci = quantile(sat_ate, probs = .025),
    upper_ci = quantile(sat_ate, probs = 0.975)
  )

ate_sat_95_17_18
```




### Lets compare the performance of different Types of Schools in 2017-2018 alongside the 2020-2021 Plot we had previously made

Trying to compare the 2020-2021 SAT to 2017 to 2018 

###Data from 2020-2021
```{r}
 

CPS_SAT_20_21_plot <- CPS_SAT_20_21|>
  select(`School Year`== "2020-2021")|>
  group_by( `Type of School`)|>
  ggplot( data = CPS_SAT_20_21, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "2020-2021 SAT Averages by Type of School",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )



CPS_SAT_20_21_plot

CPS_SAT_20_21|>
  group_by(`Type of School`)|> 
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)
```

###Data from 2017-2018
```{r}
CPS_SAT_17_18 <- CPS_SAT|>
  filter(`School Year` == "2017-2018")

CPS_SAT_17_18_plot <-CPS_SAT_17_18|>
  group_by(`Type of School`)|>
  ggplot( data = CPS_SAT_17_18, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "SAT Averages by Type of School 2017-2018 ",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )
CPS_SAT_17_18_plot

avg_sat_table_17_18<- CPS_SAT_17_18 |>
  group_by(`Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)

avg_sat_table_17_18

```
  

### When analyzing the confidence intervals for this bootstrap of the 2017-2018 year we are able to see that the upper confidence interval is 60.5 which is reflecting that the charter schools are overperforming their CPS Network counterparts by nearly 60.5 points. In the lower confidence interval we see that the lower confidence interval is -9.38 which means that it was by 9.38 points  that the CPS Network schools outperformed Charters. When comparing the results from 2017-2018 and 2020-21 we see that students from Charter Schools were outperforming CPS Network students more in 2017-2018 reflecting the 60.5 as opposed to the 15.7 in 2020-2021. We also see a difference in the lower confidence intervals as the lower confidence interval in 2017-2018 was -9.83 which is a smaller difference than the one of -46.7 reflected in 2020-2021. This shows us that on  students are underperforming in comarison to the data from 2017-2018. Could we attribute this to covid? 




###In order to try to find out if we are able to determine a correlation among the percent of college readiness and SAT scores. I will run a linear regression for both 2017-2018 school year and the 2020-2021 school year comparing the intercepts and coefficient. 

#Linear model for 2017-2018

```{r}

sat.app.17_18 <- lm(`Average Composite Score`~`Percent College Readiness`, data = CPS_SAT_17_18) 

sat.app.17_18

sat.app.17_18_graph<-sat.app.17_18|>
  ggplot( data = CPS_SAT_17_18, 
          mapping = aes(x = `Percent College Readiness`, y = `Average Composite Score`)) + 
  geom_point(aes(color = `Type of School`), na.rm = TRUE)+ geom_smooth(na.rm = TRUE) + labs(title = "2020-2021 SAT and College Readiness")

sat.app.17_18_graph

library(broom)
glance(sat.app.17_18)


```



#Linear Model for 2020-2021
```{r}
sat.app.20_21 <-lm(`Average Composite Score`~`Percent College Readiness`, data = CPS_SAT_20_21)

sat.app.20_21

sat.app.20_21_graph <- sat.app.20_21|>
  ggplot(data = CPS_SAT_20_21, 
          mapping = aes(x = `Percent College Readiness`, y = `Average Composite Score`)) +  geom_point(aes(color = `Type of School`),na.rm = TRUE) + geom_smooth(na.rm = TRUE)+ labs(title = "2020-2021 SAT and College Readiness")

sat.app.20_21_graph

library(broom)
glance(sat.app.20_21)


```

### When looking at both of these models we are able to see a poitive correlation within the Percent College Readiness and Average Composite score. As a student achieves a higher Composite score, their percent of college readiness seems to increase with it. This is true in both graphs. In the 2017-2018 Linear model we are able to see that the  y intercpet is 791.495 meaning that at 0% college readiness a student is scoring a 791.495 out of the possible 1600 in an SAT. The coefficient is 4.568 which helps us see that in this case that with every increase in 4.568 points the percent of college readiness increase by 1. In the 2020-2021 linear model we see that the y intercept is 787.949 meaning that at 0% college readiness a student is scoring a 787.949 out of the 1600. The coefficient is 4.921which helps us see that in this case with the increase in points by 4.921 we see an increase by 1. 


###CONCLUSION In this project I yearned to see if Charter schools outperform CPS Network schools. I was able to see that in actuality after analyzing the regressions  and graphs that many CPS Network schools outperform CHarter schools, meaning that they have a bigger range, while most of the times accoring to the averages, a Charter school usually outperforms a CPS Network School. We were also able to see that  when looking at the 2017-2018 school year alongside the 2020-2021 School Year there was a decline in the SAT Composite scores of students and their college readiness. Overall we must keep in mind that SAT are an exam that is arbitrary meaning that as many universities dont take the exam this does not have a true way to dictate who is college ready and who is not. SATs take a lot of time to study for and we can understand the situation of different public schools varies on their Type of institution. However this helps us see the way that we were able to correlate SAT performance, College Readiness and Type of School over time. 