---
title: "Gov 50 Final Project"
author: "Gloria Maldonado"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts


Does the influx of new populations in an area through migration and demographic changes have an impact on the amount of students enrolled in educational institutions?In this project I will look at the California Census comparing the results from the 2021 census report and the 2011 census report analyzing the differences in demographic distribution and will compare that to the populations enrolled in educational instutions at the time the census was taking. I am seeking to find out if certain demographics or ethnicities truly impact the amount of students enrolled. I hypothesize that the increasment of immigrant demographics may be a reason for which the educational enrollment may increase because it is an example of how their desires is to obtain an education.  I am interested in exploring data related to census populations in different urban areas around the country and seeing how it compares to the educational enrollment of an area. I will be looking at the variable School Enrollment specifically College or graduate school enrollment,  and the total number of races reported and their distribution within the greater population. I will look at the race variable as an explanatory variable looking at the changes in demographic over the years. My outcome variable will be the educational enrollment in college or grad school. For example, if I observe an increase in the presence of the "non-white hispanic" population, when comparing 2011- 2021,  and likewise see an increase in the enrollment of "non-white hispanic" population from 2011 to 2021, I can attribute that this positive change can be attributed to the demographic's growing desire to obtain an education. If this was true it would be something to help supporty my hypothesis that certain demographics aligning with their migration, have desires to gain educational attainment seen through their enrollment in higher education institutions. On the other hand, if I observe that there is a decrease from 2011-2021 in the educational enrollement of "Non-white hispanic" yet an increase in the demographic percentage of "non-white hispanic" this would be a negative relationship and would prove against my hypothesis. 



	
	Variable Name	Description
"School 
and Test"	School Year	The school year in which the tests were taken
	School ID	Six digit school identification number
	School Name	School's short name
	Test	Assessment (i.e., PSAT9, PSAT10, SAT)
Attainment	# Students	Number of students taking the PSAT or SAT assessment who received a Total Score. For information on inclusion and exclusion rules, see below.
	Average Composite Score 	The average composite score for the students taking the test.
	# Students with EBRW Score	Number of students taking the PSAT or SAT assessment who received an EBRW Score. For information on inclusion and exclusion rules, see below.
	Average EBRW Score	The average Evidence-Based Reading and Writing score for the students taking the  Spring PSAT or SAT assessment.
	# Students with Math Score	Number of students taking the PSAT or SAT assessment who received a Math Score. For information on inclusion and exclusion rules, see below.
	Average Math Score	The average Math score for the students taking the Spring PSAT or SAT assessment.
	% Meeting College Readiness Benchmark	The percentage of students taking the Spring PSAT or SAT assessment who scored at or above the grade-appropriate combined college readiness benchmark.
	% Partially Met State Standards	The percentage of students who only partially met standards and demonstrated minimal understanding on the PSAT or SAT.
	% Approached State Standards	The percentage of students who were approaching the proficiency level and demonstrated incomplete understanding on the PSAT or SAT.
	% Met State Standards	The percentage of students who met the proficiency level and demonstrated adequate understanding on the PSAT or SAT.
	% Exceeded State Standards	The percentage of students who exceeded the proficiency level and demonstrated a thorough understanding on the PSAT or SAT.
	% Met or Exceeded State Standards	The percentage of students who met or exceeded the proficiency level set by the Illinois State Board of Education for students on the PSAT or SAT.



```{r}
library(tidyverse)
library(tidyselect)
library(dplyr)


library(readr)
new_sat_data <- read_csv("~/Downloads/Gov 50/new_sat_data.csv")
View(new_sat_data)

library(readr)
type_school_sat <- read_csv("~/Downloads/Gov 50/type_school_sat.csv")
View(type_school_sat)
``` 

fixing up sat data 

```{r}


new_join_sat <- inner_join(new_sat_data, type_school_sat, by = "School ID" )

CPS_SAT<- new_join_sat

CPS_SAT<- CPS_SAT|>
  rename("Percent College Readiness" = contains( "Readiness", ignore.case = TRUE, vars = NULL))|>
  filter(Test == "SAT")


# Making a data set for 2020-2021 School Year 

CPS_SAT_20_21 <- CPS_SAT|> 
  filter(`School Year` == "2020-2021")

#investigating based on 2020-2021 School Year 

CPS_SAT_20_21_plot <- CPS_SAT_20_21|>
  select(`School Year`== "2020-2021")|>
  group_by( `Type of School`)|>
  ggplot( data = CPS_SAT_20_21, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "2020-2021 SAT Averages by Type of School",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )

CPS_SAT_20_21_plot

```

``` {r}
#Average of the `Average Composite Scores` by `Type of School`

avg_sat_table_20_21<- CPS_SAT_20_21 |>
  group_by(`Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)

avg_sat_table_20_21


#Finding out the difference in averages between Charters and CPS Network 
# this is the differenc between the SAT SCORE OF CPS and a Charter 
sat_ate <- CPS_SAT_20_21|>
  group_by(`Type of School`)|> 
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`) |> 
  mutate(sat_ate = `Charter` - `CPS Network`)|>
  pull(sat_ate)

sat_ate
```


```{r} 

college_readiness_average <- CPS_SAT_20_21|> 
  group_by(`Type of School`)|>
  summarise(college_readiness_avg = mean(`Percent College Readiness` , na.rm = TRUE))

college_readiness_average

#We will see if there is an average treatement effect for going to charter 

college_readiness_ate<- CPS_SAT_20_21 |>
  group_by(`Type of School`)|>
  summarise(college_readiness_avg = mean(`Percent College Readiness` , na.rm = TRUE))|>
  pivot_wider(values_from = college_readiness_avg, names_from = `Type of School`)|>
  mutate(college_readiness_ate = `Charter` - `CPS Network`)|>
  pull(college_readiness_ate)

college_readiness_ate

```



creating bootstraps to find out the various quantiles of the averages of different schools that fall in every different type  


20-21 
```{r}
library(infer)
set.seed(02138) 

ate_boots_20_21<- CPS_SAT_20_21|>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000)|>
  group_by(replicate, `Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg , names_from = `Type of School`)|>
  mutate(sat_ate = `Charter` - `CPS Network`)

ate_boots_20_21 


ate_sat_95_20_21 <- ate_boots_20_21|>
  ungroup(replicate)|>
  summarize(
    lower_ci = quantile(sat_ate, probs = .025),
    upper_ci = quantile(sat_ate, probs = 0.975)
  )

ate_sat_95_20_21

```




```{r}
library(infer)
set.seed(02138) 

ate_boots_17_18<- CPS_SAT_17_18|>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000)|>
  group_by(replicate, `Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg , names_from = `Type of School`)|>
  mutate(sat_ate = `Charter` - `CPS Network`)

ate_boots_17_18


ate_sat_95_17_18 <- ate_boots_17_18|>
  ungroup(replicate)|>
  summarize(
    lower_ci = quantile(sat_ate, probs = .025),
    upper_ci = quantile(sat_ate, probs = 0.975)
  )

ate_sat_95_17_18
```



Math lets look how that impacts 

```{r}
math_CPS <- CPS_SAT_20_21|>
  group_by(`Type of School`)|>
  summarise(math_avg = mean(`Average Math Score`, na.rm = TRUE))|>
  pivot_wider(values_from = math_avg, names_from = `Type of School`) 

math_CPS

```





correlation 

try to find the correlation of the percentage that are college ready compared to the 

i dont think this is going to be accurate becaus ein the past the two variables were calculates in a way where they were on the same scale and percentage level. 

```{r}
corr_sat_college_readiness <- cor(CPS_SAT$`Average Composite Score`, CPS_SAT$`Percent College Readiness`, method = "pearson")

corr_sat_college_readiness
```


Trying to compare the 2020-2021 SAT to 2017 to 2018 
```{r}


# first I could filter so its only 2017-2018 and 2020-2021 but it may be interesting to look at alll the years 

CPS_SAT_20_21_plot <- CPS_SAT_20_21|>
  select(`School Year`== "2020-2021")|>
  group_by( `Type of School`)|>
  ggplot( data = CPS_SAT_20_21, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "2020-2021 SAT Averages by Type of School",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )

CPS_SAT_20_21_plot

CPS_SAT_17_18 <- CPS_SAT|>
  filter(`School Year` == "2017-2018")

CPS_SAT_17_18_plot <-CPS_SAT_17_18|>
  group_by(`Type of School`)|>
  ggplot( data = CPS_SAT_17_18, 
          mapping = aes(x = `Type of School`, y = `Average Composite Score`)) + 
  geom_boxplot(na.rm = TRUE) + 
  labs( title = "SAT Averages by Type of School 2017-2018 ",
    x = "Type of School",
    y = "Average SAT Composite Score" 
  )
CPS_SAT_17_18_plot

avg_sat_table_17_18<- CPS_SAT_17_18 |>
  group_by(`Type of School`)|>
  summarise(sat_avg = mean(`Average Composite Score`, na.rm = TRUE))|>
  pivot_wider(values_from = sat_avg, names_from = `Type of School`)

avg_sat_table_17_18

CPS_17_20 <- CPS_SAT|>
  filter(`School Year` == c("2017-2018" ,  "2020-2021"))

CPS_17_20
   
```
  


corr_sat <- cor.test(CPS_SAT_17_18$`Average Composite Score`, CPS_SAT_20_21$`Average Composite Score`, method = "pearson", use= "complete.obs")

corr_sat





linear modeel 

#2017 and 2018 

```{r}

sat.app.17_18 <- lm(`Average Composite Score`~`Percent College Readiness`, data = CPS_SAT_17_18) 

sat.app.17_18

sat.app.17_18_graph<-sat.app.17_18|>
  ggplot( data = CPS_SAT_17_18, 
          mapping = aes(x = `Percent College Readiness`, y = `Average Composite Score`)) + 
  geom_point(aes(color = `Type of School`), na.rm = TRUE)+ geom_smooth(na.rm = TRUE) + labs(title = "2020-2021 SAT and College Readiness")

sat.app.17_18_graph

library(broom)
glance(sat.app.17_18)

fit.x
```



#2020-2021
```{r}
sat.app.20_21 <-lm(`Average Composite Score`~`Percent College Readiness`, data = CPS_SAT_20_21)

sat.app.20_21

sat.app.20_21_graph <- sat.app.20_21|>
  ggplot(data = CPS_SAT_20_21, 
          mapping = aes(x = `Percent College Readiness`, y = `Average Composite Score`)) +  geom_point(aes(color = `Type of School`),na.rm = TRUE) + geom_smooth(na.rm = TRUE)+ labs(title = "2020-2021 SAT and College Readiness")

sat.app.20_21_graph

library(broom)
glance(sat.app.20_21)


```



